---
title: "BSTA 6100 Lab 1 Report"
author: "Your Name"
format:
  html:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: lumen
    css: assets/styles.css
    self-contained: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Complete the following exercises in R / RStudio. It is important that you keep your answers inside the ".answer" blocks. Do NOT delete the three colon blocks (:::) and keep your responses between them. Use the tutorial for help, and don't be afraid to ask questions.

**Collaborators:** If applicable, replace this text with the names of anyone you collaborated with on this project.

You can find a list of colors you can use in R [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf). You can also directly use RGB colors via the `rgb()` function (see the help for more), or hex codes. If you don't know what those are, stick to the named colors in the linked PDF.

## Background

Sleep apnea, literally “without breath”, occurs during sleep when one’s breath becomes very shallow or briefly stops altogether.  Episodes of sleep apnea typically last 20 to 40 seconds and can happen many times a night in some people. Sleep apnea is of various types, one of which is obstructive sleep apnea (OSA) and is the focus of this dataset.  OSA has physical causes; it occurs when something partly or completely blocks the upper airway during sleep.  OSA is usually associated with a reduction in blood oxygen saturation.   Notably, approximately 30% of the general population suffer from OSA.

Obstructive sleep apnea (OSA) is characterized by specific events and their severity.  OSA events are of two types: an apnea, characterized by complete cessation of airflow for at least 10 seconds, or a hypopnea in which airflow decreases by 50 percent for 10 seconds or decreases by 30 percent if there is an associated decrease in the oxygen saturation or an arousal from sleep. Sleep apnea severity is the frequency of event occurrence per hour and is assessed using the apnea-hypopnea index (AHI). An AHI of less than 5 is considered normal. An AHI of 5-14 is mild; 15-29 is moderate and 30 or more events per hour characterizes severe sleep apnea.

As a result of OSA, nocturnal oxygenation levels are often lower than normal. Two common measures are the percentage of total sleep time spent at Sao2 less than 90% and the minimum nocturnal Sao2 listed in polysomnography reports. OSA is of concern because of its relationship to cardiovascular disease, with a diagnosis of moderate-to-severe disease (AHI ≥ 15) being identified as an independent risk factor for all-cause and cardiovascular mortality. Patients suffering from OSA present with a chronic enhancement in sympathetic adrenergic activity (system involved in maintaining homeostasis) that is considered one of the major mechanisms in the development of cardiovascular health issues in this population. Such hemodynamic instability is hypothesized to increase the risk for intraoperative and/or postoperative morbidity in OSA patients undergoing surgery.

# Lab Assignment

In this assignment, we'll be using a data set we will call `hypoxia` that contains data from a study of intraoperative mean arterial pressure in obstructive sleep apnea patients undergoing weight loss (bariatric) surgical procedures.  Participants are 281 adult patients diagnosed with obstructive sleep apnea within two years prior to their bariatric surgery. In addition to mean arterial pressure, the dataset includes information on demographics, nocturnal oxygen status, comorbidities, type and duration of surgery, vasopressors, and heart rate. The data are cleaned and does contain some missing data. The source of these data is the study by Turan et al. “Relationship between Chronic Intermittent Hypoxia and Intraoperative Mean Arterial Pressure in Obstructive Sleep Apnea Patients Having Laparoscopic Bariatric Surgery”. *Anesthesiology* 2015; 122: 64-71.

(Note: study text, background, etc., is taken from https://causeweb.org/tshs/hypoxia/). A data dictionary is available in [Hypoxia Data Dictionary](Hypoxia Data Dictionary.pdf)

## Question 1

::: {.question}
We'll start by reading in the data. The data are stored in a file called `hypoxia.csv`. You'll need to give the name of this file (*in quotes!*) to `read.csv()`. When you read it in, call the data set `hypoxia`. Make sure you set `stringsAsFactors = TRUE`. 
:::

::: {.answer}
```{r}
#| label: question1
#| error: true
# Replace this comment with code required for Question 1. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```
:::

## Question 2

:::{.question}
Make a boxplot of the variable that represents minimum nocturnal oxyhemoglobin saturation. Be sure to give your boxplot an appropriate title, axis labels, etc. Based on this boxplot, does it appear that the data are symmetric?
:::

:::{.answer}
```{r}
#| label: question2
#| error: true

# Replace this comment with code required for Question 2. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

Replace this text with your answers to Question 2.

:::

## Question 3

:::{.question}
Make a histogram of the same variable from Question 2, being sure to include a proper title and axis labels. Play around with the `breaks` argument until you're satisfied with your histogram. Was your prediction about the symmetry of the data in Question 2 accurate? In the future, do you think it's always appropriate to describe shape based on a boxplot alone?
:::

:::{.answer}
```{r}
#| label: question3
#| error: true
# Replace this comment with code required for Question 4. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

Replace this text with your answers to Question 3.

:::

## Question 4

:::{.question}
Make a side-by-side boxplot of time spent during sleep with oxyhemoglobin saturation less than 90% by smoking status. As always, be sure to provide an appropriate title and axis labels. In 1-2 sentences, describe what you see.
:::

:::{.answer}
```{r}
#| label: question4
#| error: true
# Replace this comment with code required for Question 4. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

Replace this text with your answer to Question 4.

:::

## Question 5

:::{.question}
Is worse sleep apnea associated with lower arterial blood pressure during bariatric surgery? Make a scatterplot of mean arterial pressure during surgery vs. time spent asleep with blood oxygenation below 90% (remember we say "y vs. x", so MAP should be on the y axis). Write 1-2 sentences describing what you see.
:::

:::{.answer}
```{r}
#| label: question5
#| error: true
# Replace this comment with code required for Question 5. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

Replace this text with your 1-2 sentence answer to Question 5.

:::

## Question 6
:::{.question}
Remake the scatterplot you created in Question 5, this time coloring the points by smoking status. 

**Note:** Because `Smoking` is a numeric variable coded 0/1, you'll need to either color by `factor(hypoxia$Smoking)` or by `hypoxia$Smoking + 1`. 
:::

:::{.answer}
```{r}
#| label: question6
#| error: true
# Replace this comment with code required for Question 6. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```
:::


## Question 7
:::{.question}
Is worse sleep apnea associated with lower arterial blood pressure during bariatric surgery? Fit a linear regression model with outcome variable MAP and covariate amount of time spent sleeping with oxyhemoglobin concentration below 90%. Write the equation of the fitted regression line.
:::

:::{.answer}
```{r}
#| label: question7
#| error: true
# Replace this comment with code required for Question 7. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

The fitted regression line is $\hat{Y} = $

:::

## Question 8

:::{.question}
This is a preview of more exploration of linear regression to come! An important *diagnostic* for a linear regression model is a plot of residuals vs. the fitted values (i.e., the $\hat{Y}$'s). This helps us check

1. the assumption that the residuals have mean zero, which tells us that the model is correctly specified, and
2. the assumption that the residuals have a constant variance $\sigma^2$. 

To access the residuals of a model object, use `$residuals`. So if your model is stored as `mod1`, for example, you can get the residuals with `mod1$residuals`. Similarly for fitted values: `mod1$fitted.values`.

Create a plot of the residuals vs. fitted values for the model you fit in Question 7, and discuss what you see. You don't need to think about how the assumptions come into play, just describe the scatterplot.
:::

:::{.answer}

```{r}
#| label: question8
#| error: true

# Replace this comment with code required for Question 8. (Remember that this text is a comment, so R ignores it; you can delete it if you want.)

```

Replace this text with your 1-2 sentence answer to Question 8.
:::


# Wrap-Up and Submission

At the top of the document, make sure you've changed the `author` field to your name (in quotes!).

When you've finished the lab, click the **render** button one last time, and submit the compiled HTML file to Canvas.
